Visual Question Answering with DeepProbLog

### 2022-03-20 paper/paper.tex
Write further

\begin{abstract}
	This report focusses on Visual Question Answering, where a Neuro Symbolic AI approach with a knowledge base is compared with a purely neural network based approach. From the experiments, it follows that DeepProbLog, the framework used for the Neuro Symbolic AI approach, is able to achieve the same accuracy as the pure neural network based approach with almost 200 times less iterations. Clearly, the training is much more targeted, but however, comes at a cost. The algebraic operators internal to DeepProbLog are extremely costly and hence the actual training time is more then twenty times slower, when both approaches were trained on a CPU. Another drawback of DeepProbLog is that no easy speedups can be achieved, since the algebraic operators only work on CPU's, and hence cannot benefit from accelerators such as GPU's.
\end{abstract}

@@

\section{ORGANIZATION OF THE PAPER}
This paper will first provide some necessary background in Section \ref{sec:literature_survey} on certain types of datasets that are typically used for Visual Question Answering purposes. Then, Section \ref{sec:approach} will dive deeper in the different components constituting the overall system. Section \ref{sec:experiments}, furthermore, provides some results of conducted experiments. A main focus in this paper is to outline the advantages of using a Neuro-Symbolic AI approach (offered by the DeepProbLog framework), instead of a purely neural network based approach. These experiments are listed in Section \ref{sec:experiments}. Finally, the main findings of this paper are repeated in Section \ref{sec:conclusions}.

@@

Clearly, both system 1 and system 2 are actively used when answering these questions. One could wonder if neural networks alone could answer these questions without having an explicit system 2 encoding (i.e. the rule based knowledge base). Intuitively, it makes sense that if certain facts of the world are known\footnote{Facts can be encoded, e.g. counting the number of spheres is simply a matter of detecting all the spheres in the image, after which a mathematical summation is a statement in the knowledge base.}, learning can proceed much more quickly\footnote{Not to say that learning might even be impossible if a lot of background knowledge is required.}. Seen from an optimization viewpoint, errors made during prediction in this setup can be targeted exactly, which makes the optimization process more targeted as well, and hence more efficient. Finally, this paper also provides evidence for these statements, since in subsection \ref{subsec:experiments_NN_vs_deepproblog}, the comparison between a VQA implementation with DeepProbLog is made with a purely CNN based approach.

@@

A. Santoro et al. conducted similar experiment with the Sort-of-CLEVR dataset as in this paper, where they were able to achieve an accuracy of 63\% on relational questions with CNN's, whereas an accuracy of 94\% for both relational and non-relational questions was achieved with CNN's complemented with an RN. For them, augmenting the model with a relational module, such as an RN, turned out to be sufficient to overcome the hurdle of solving relational questions \cite{sort_of_clevr_dataset}.

@@

These images are accompanied with a question about a random object, which can be one of the following:
\begin{itemize}
    \item Non-relational - What is the shape of this object \footnote{I.e. the shape will be either a square or a circle.}?
    \item Non-relational - Is this object located on the left side of the image?
    \item Non-relational - Is this object located on the bottom side of the image?
    \item Binary - How many objects have the same shape as this object?
\end{itemize}

These questions are encoded in a one-hot encoding, after which they are stored in a CSV file, along with the expected answers. To make the training process more efficient, a training and test dataset has been generated beforehand.

@@

\subsection{COMPARISONS WITH PURE SYSTEM 1 APPROACHES}
\label{subsec:experiments_NN_vs_deepproblog}
\subsubsection{Experiment: 2x2 grid}
The loss curves of both the DeepProbLog approach, as well as the purely neural network based approach, are visualized respectively in Figure \ref{fig:loss_curve_deepproblog_2x2} and Figure \ref{fig:loss_curve_pure_NN_2x2}.

@@

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{loss_curve_pure_NN_2x2.png} 
    \captionsetup{justification=centering}
    \caption{Pure NN based approach 2x2: Loss curve}
    \label{fig:loss_curve_pure_NN_2x2}
    \end{center}
\end{figure}
An extremely important remark to be made is the difference between `Number of iterations' and `Number of epochs'. By the number of iterations, the number of forward and backward passes of a batch (with size 32) is meant, whereas the number of epochs denote the number of times all the images of the training set are forward and backwardly passed. In this application, a training size of 10 000 was used, hence one epoch consists of 312.5 iterations.

@@

\subsubsection{Experiment: 6x6 grid}
The loss curves for the 6x6 experiment of DeepProbLog and the neural network based approach are depicted in respectively Figure \ref{fig:loss_curve_deepproblog_6x6} and Figure \ref{fig:loss_curve_pure_NN_6x6}.

@@

In these experiments, the training time of the purely neural network approach was 20 minutes\footnote{For fair speed estimation, training has been conducted on the CPU.}, while the training time of the DeepProbLog approach was a little under eight hours\footnote{However, it should be mentioned that approximately half of the time was spent on calculating the accuracy each time, since the whole testing training set had to be forwarded through the network, while for the training iterations this is only one batch.}.

The most important observation here is that the purely neural network approach overfits quickly. Another important remark is the fact that both approaches are clearly not longer able to attain 100\% accuracy.

@@

\section{CONCLUSIONS}
\label{sec:conclusions}
The strengths of the Neuro Symbolic AI field have been demonstrated in the context of Visual Question Answering. By using DeepProbLog, the chosen framework for the Neuro Symbolic AI task, it became clear that almost 200 times less iterations are required to achieve the same accuracy as a purely neural network based approach. However, due to the costly algebraic operators, the total training time of the DeepProbLog approach was still more than twenty times slower, compared to the purely neural network based approach.

>>>>

\begin{abstract}
	This report focusses on Visual Question Answering, where a Neuro Symbolic AI approach with a knowledge base is compared with a purely neural network based approach. From the experiments, it follows that DeepProbLog, the framework used for the Neuro Symbolic AI approach, is able to achieve the same accuracy as the pure neural network based approach with almost 200 times less iterations. Clearly, the training is much more targeted, but however, comes at a cost. The algebraic operators internal to DeepProbLog are extremely costly and hence the actual training time is considerably slower. Another drawback of DeepProbLog is that no easy speedups can be achieved, since the algebraic operators only work on CPU's, and hence cannot benefit from accelerators such as GPU's.
\end{abstract}

@@

\section{ORGANIZATION OF THE PAPER}
This paper will first provide some necessary background in Section \ref{sec:literature_survey} on certain types of datasets that are typically used for Visual Question Answering purposes, along with introducing some results of other researchers. Then, Section \ref{sec:approach} will dive deeper in the different components constituting the overall system. Section \ref{sec:experiments}, furthermore, provides some results of conducted experiments. A main focus in this paper is to outline the advantages of using a Neuro-Symbolic AI approach (offered by the DeepProbLog framework), instead of a purely neural network based approach. These experiments are listed in Section \ref{sec:experiments}. Finally, the main findings of this paper are repeated in Section \ref{sec:conclusions}.

@@

Clearly, both system 1 and system 2 are actively used when answering these questions. One could wonder if neural networks alone could answer these questions without having an explicit system 2 encoding (i.e. the rule based knowledge base). Intuitively, it makes sense that if certain facts of the world are known\footnote{Facts can be encoded, e.g. counting the number of spheres is simply a matter of detecting all the spheres in the image, after which a mathematical summation is a statement in the knowledge base.}, learning can proceed much more quickly\footnote{Not to say that learning might even be impossible if a lot of background knowledge is required.}. Seen from an optimization viewpoint, errors made during prediction in this setup can be targeted exactly, which makes the optimization process more targeted as well, and hence more efficient. Finally, this paper also provides evidence for these statements, since in subsection \ref{subsec:experiments_NN_vs_deepproblog}, the comparison between a VQA implementation with DeepProbLog is made with a purely neural network based approach.

@@

A. Santoro et al. conducted similar experiment with the Sort-of-CLEVR dataset as in this paper, where they were able to achieve an accuracy of 63\% on relational questions with CNN's. In contrast, an accuracy of 94\% for both relational and non-relational questions was achieved with CNN's complemented with an RN. For them, augmenting the model with a relational module, such as an RN, turned out to be sufficient to overcome the hurdle of solving relational questions \cite{sort_of_clevr_dataset}.

@@

These images are accompanied with a question about a random object, which can be one of the following:
\begin{itemize}
    \item Non-relational - What is the shape of this object \footnote{The shape will be either a square or a circle.}?
    \item Non-relational - Is this object located on the left side of the image?
    \item Non-relational - Is this object located on the bottom side of the image?
    \item Relational - How many objects have the same shape as this object?
\end{itemize}

These questions are encoded in a vector encoding, after which they are stored in a CSV file, along with the expected answers. A training and test dataset has been generated beforehand, in order to make the training process more efficient.

@@

\subsection{COMPARISONS WITH PURE SYSTEM 1 APPROACHES}
\label{subsec:experiments_NN_vs_deepproblog}
\subsubsection{Experiment - 2x2 grid}
The loss curves of both the DeepProbLog approach, as well as the purely neural network based approach, are visualized respectively in Figure \ref{fig:loss_curve_deepproblog_2x2} and Figure \ref{fig:loss_curve_pure_NN_2x2}.

@@

\begin{figure}[htp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{loss_curve_pure_NN_2x2.png} 
    \captionsetup{justification=centering}
    \caption{Pure NN based approach 2x2: Loss curve}
    \label{fig:loss_curve_pure_NN_2x2}
    \end{center}
\end{figure}
An extremely important remark to be made is the difference between `Number of iterations' and `Number of epochs'. By the number of iterations, the number of forward and backward passes of a batch (with size 32) is meant, whereas the number of epochs denote the number of times all the images of the training set are forward and backwardly passed. In this application, a training size of 10 000 was used, hence one epoch consists of 312.5 iterations.

@@

\subsubsection{Experiment - 6x6 grid}
The loss curves for the 6x6 experiment of DeepProbLog and the neural network based approach are depicted in respectively Figure \ref{fig:loss_curve_deepproblog_6x6} and Figure \ref{fig:loss_curve_pure_NN_6x6}.

@@

In these experiments, the training time of the purely neural network approach was 20 minutes\footnote{For fair speed estimation, training has been conducted on the CPU.}, while the training time of the DeepProbLog approach was a little under eight hours\footnote{However, it should be mentioned that approximately half of the time was spent on calculating the accuracy each time, since the whole testing testing set had to be forwarded through the network, while for the training iterations this is only one batch.}.

The most important observation here is that the purely neural network approach overfits quickly and also achieves a considerably lower accuracy (68\% instead of almost 75\% for DeepProbLog). Another important remark is the fact that both approaches are clearly not longer able to attain 100\% accuracy. However, if the DeepProbLog network could train longer, it would be able to converge even further.

@@

\section{CONCLUSIONS}
\label{sec:conclusions}
The strengths of the Neuro Symbolic AI field have been demonstrated in the context of Visual Question Answering. By using DeepProbLog, the chosen framework for the Neuro Symbolic AI task, it became clear that almost 200 times less iterations are required to achieve the same accuracy as a purely neural network based approach. However, due to the costly algebraic operators, the total training time of the DeepProbLog approach was considerably slower, compared to the purely neural network based approach.
